---
title: "PSTAT 5LS: Homework 05 -- \\textcolor{blue}{\\textbf{SOLUTIONS}}"
subtitle: "Question 7: Say It with Confidence!"
author: "Your Name Here"
date: "Spring 2026"
output: pdf_document
urlcolor: blue
header-includes:
  - \usepackage{mdframed}
toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(fig.width = 6, fig.height = 3.5, fig.align = 'center')
library(pstat5lsSBI)
set.seed(342)  ## DO NOT CHANGE THIS VALUE FOR ANY REASON!
```

**VERY IMPORTANT:** Remember to start by making a copy of this document, and only edit the copy - **leave this original file untouched** (so your files don't get overwritten every time the server updates). [Click here for instructions on how to do that.](https://docs.google.com/document/d/12SKM9YPDtStpjVptOvMjVIGZjQcaJFIa_DhJMXE4mMQ/edit?usp=sharing)

Please note that this is only **one question** of Homework 5; the remainder of the questions appear on Gradescope in an "Online Assignment" format. (Please see Canvas for further details.)

---

## Part I: Simulating a Normal Distribution

We have previously seen how simulations can be a very powerful tool in a statistician's arsenal. To that end, did you know that we can actually simulate values from a normal distribution? The function `rnorm()` does exactly that; specifically, the command
```{r, eval = F}
rnorm(n, mean = mu, sd = sigma)
```
will generate `n` values from a $N(\mu, \sigma)$ distribution. As a reminder, this is a _different_ function from the `pnorm()` and `qnorm()` functions you've seen in Lab 04. Recall:

-   `pnorm()` gives us **p**robabilities (hence the `p`)
-   `qnorm()` gives us percentiles (aka **q**uantiles; hence the `q`)
-   `rnorm()` gives us **r**andom draws (hence the `r`)


### \textcolor{blue}{\textbf{Question 1}}: Generate a list of `n` = 1000 numbers from a _N_(3, 1.2) distribution, and assign the result to a variable called `random_normal_sample`. Then, generate a histogram of these 1000 values and check that the histogram looks like a normal distribution (bell-shaped, symmetric, centered at the appropriate value).

```{r question1}
## Replace this line with your answers; don't forget to delete the double hashtags!
```

--- 

## Part II: Confidence Intervals

Recall that our formula for the confidence interval for a population mean $\mu$ is 
$$ \bar{x} \pm t^{\ast} \cdot \frac{s}{\sqrt{n}} $$
where $\bar{x}$ denotes the sample mean, $t^{\ast}$ is the **critical value**, $s$ is the sample standard deviation, and $n$ is the sample size. Further recall (from lecture and lab) that the critical value $t^{\ast}$ can be found using the `qt()` function, using $(n - 1)$ degrees of freedom.


### \textcolor{blue}{\textbf{Question 2}}: Pretend we didn't know the true mean $\mu$ of the distribution from which our `random_normal_sample` values were sampled. Construct a 95\% confidence interval for $\mu$ using the `random_normal_sample` values. Assign the lower endpoint of your interval to a variable called `ci_lower`, and your upper endpoint to a variable called `ci_upper`.

**Some relevant `R` functions:** `mean()`, `qt()`, `sd()`, and `sqrt()`.

```{r question2}
## Replace this line with your answers; don't forget to delete the double hashtags!
```

If you did your variable assignments correctly, the code below should display your interval:
```{r}
c(ci_lower, ci_upper)
```

---

Because our `random_normal_sample` values were just a simulation, we can do something we wouldn't be able to do with non-simulated data: compare against the truth! Specifically, we actually know that $\mu = 3$, since we generated data from a _N_(3, 1.2) distribution. Again, in a real-world setting, we wouldn't know the value of $\mu$ - we only know $\mu$ in this case because we conducted a simulation. 


### \textcolor{blue}{\textbf{Question 3}}: Is the true value of $\mu = 3$ contained in the interval you constructed for Question 2? Is this a problem?

\begin{mdframed}

Add your answers here.

\end{mdframed} 

---


## Closing Instructions

Congratulations on finishing! When you are done, please do the following:

1)    Click the "Knit" button at the top of your RStudio screen (it should have an icon that looks like a ball of yarn). This will create a PDF out of your work.

2)    Check that all formatting appears correctly, and that all equations are rendering properly. If not, go back and fix any issues.

3)    Download your PDF.

4)    Upload your PDF to the appropriate part of the Homework 5 assignment in Gradescope.