---
title: "PSTAT 5LS Lab 01"
author: "Professor Jack Miller, and Ethan Marzban"
date: "Spring 2026"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(fontawesome)
```


## Learning Objectives

### `R` Learning Objectives

1)  Learn the difference between R, RStudio, and R Markdown
2)  Learn how to use R to do basic calculations
3)  Understand how R functions work
4)  Learn how to compute some basic statistical summaries in R


### Statistical Learning Objectives

1)  Understand that technology can be helpful in statistics


## Getting Sta-`R`-Ted

### What is `R`?

In Statistics, we often use computers to analyze data. There are a lot of
"statistical computing environments" that can help you do statistical
analyses. One of the most popular (and powerful) is called R. R is a
programming language that is designed for manipulating data, doing
calculations, and making graphical displays. R works by writing **R code**.

That might sound scary, but don't worry: this is not a programming class.
Over the course of the quarter, you'll learn how to edit and write some basic
R code to help you analyze data to answer research questions. Our goal
with PSTAT 5LS labs is to help you learn the basics of R and R coding, but
through the lens of answering statistical questions.

There are a lot of "R" words floating around. What's going on?

-   _R_ is a "statistical computing environment" that's designed for manipulating data, generating plots, and performing analyses. It's also a programming language. You'll be using R this quarter.
-   _RStudio_ is an "integrated development environment (IDE)" for R (you'll never have to hear the term IDE again in this class). Basically, RStudio is a pretty interface that makes working with R easier. You
use R inside of RStudio. If R is ice cream, RStudio is the cone or cup.
-   _R Markdown_ is a way to write pretty analysis reports that combines R
code, R output (plots, analysis results, etc.) and text in one document. This lab document is an R Markdown document!


### R Markdown

R Markdown lets you combine text, R code, and plots in one pretty, reproducible report. If you're curious about this, you can find more details on using R Markdown at http://rmarkdown.rstudio.com.

You'll explore an R Markdown document in greater detail in Question 7 of Homework 1.

<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt1

Periodically, we will ask you to run various pieces of code or modify various code chunks in this lab document. Here is our first one!

```{r tryIt1, exercise = TRUE, error = T}
# Use this code chunk to print the phrase "Hello World!" to the screen
print("Hello world!")
```
</div>


## Becoming a Code-`R`

### Using `R` as a Calculator

At its most basic, R is a fancy calculator. Let's try this out!

<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt2
```{r tryIt2, exercise = TRUE, error = T}
# Use this code chunk to compute what 5 times 7 is (we use a star * for multiplication)


```

```{r tryIt2-solution}
5 * 7
```
</div>


When you run the chunk, you'll see a `[1]` before the output of 35. Just ignore this. The result is 35.

All of the symbols you think you'd use to do math work: `+` for addition, `-` for subtraction, `*` for multiplication, `/` for division, `ˆ` for exponentiation.


<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt3

Use the code chunk below to calculate the value of $2^{10}$
```{r tryIt3, exercise = TRUE, error = T}
## Replace this line with your code
```

```{r tryIt3-solution}
2^10
```
</div>

### `R` Assginment Syntax

Often, we want `R` to remember the results of a calculation so that we can use it later. We can give the result of some code a name by **assigning** it to something.

```{r}
x <- 36 / 6
```

We read that code as "x gets 36 / 6". The arrow is made using the less-than symbol (`<`, shift + comma on a US English keyboard) and a hyphen. 

<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt4

Before running the chunk below, try and guess what it will return. Then, run the chunk and see if you were correct!
```{r tryIt4, exercise = TRUE, error = T}
x <- 36 / 6
x
```
</div>

One important piece of information to keep in mind is that `R` is **case-sensitive.** For example, `x` and `X` are treated as two separate variables.

<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt5

Before running the chunk below, try and guess what it will return. Then, run the chunk and see if you were correct!
```{r tryIt5, exercise = TRUE, error = T}
x <- 36 / 6
X
```
</div>

### `R` Errors

`R` errors are actually quite informative. For example, notice that when we ran `X` after assigning a value to `x`, `R` said "Error: object 'X' not found". That is, it is quite literally telling us that it doesn't know what `X` means - because we forgot to define it! 

There are a few other types of errors `R` returns. Run the following cells to explore each of them.

```{r badname1, exercise = TRUE, error = T}
tik tok <- 12
```

```{r badname2, exercise = TRUE, error = T}
4eva <- 4 * 2
```

```{r badname3, exercise = TRUE, error = T}
_hi_mom <- 5^2
```

The errors saying "unexpected symbols" or "unexpected input" are R's way of telling you that these names are not allowed, and that you should use a different name. Here's how we’d correct these:

```{r goodname1, exercise = TRUE, error = T}
tiktok <- 12
```

```{r goodname2, exercise = TRUE, error = T}
forever <- 4 * 2
```

```{r goodname3, exercise = TRUE, error = T}
hi_mom <- 5^2
```

<div style="background-color:#eaf0e4">
### `r fontawesome::fa("lightbulb")` ProTip
If you want to simultaneously perform a variable assignment and also display the object being assigned to a variable, enclose the entire variable assignment statement in parentheses:

```{r}
(y <- 2 + 3)
```
</div>

## Functions and Statistics in `R`

We’ll be using R primarily through the use of functions. A function generally looks something like this:

```{r eval = FALSE}
function.name(argument1, argument2, ...)
```

Notice that we have the name of the function, followed immediately by an open parenthesis ‘(‘ without a space, then a sequence of arguments to the function, separated by commas ‘,‘, then a close parenthesis ')'. 

<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt6

The function `cos()` computes the cosine of its input. Use the code chunk below to calculate the value of $3^{\cos(4)}$, and assign it to a variable called `three_cos_four`

```{r tryIt6, exercise = TRUE, error = T}
## Replace this line with your code
```

```{r tryIt6-solution}
three_cos_four <- 3 ^ cos(4)
```
</div>

One of the many neat things about R is that it has functions that can help us perform statistical analyses and produce graphical summaries!

Toward the end of Slide Set 1, we were exposed to some common measures of center: the **mean** and the **median**. Sure enough, R has functions that allow us to calculate both of these!

-   `mean()` is used to calculate the mean of a list of numbers
-   `median()` is used to calculate the median of a list of numbers

<div style="background-color:#ede7d8">
### `r fontawesome::fa("triangle-exclamation")` Caution
In R, lists of numbers need to be inputted between a pair of parentheses with a "c" in front. So, for example,
```{r, eval = F}
c(1, 2, 3)
```
is a valid list of numbers, whereas
```{r, eval = F}
1, 2, 3
```
and
```{r, eval = F}
(1, 2, 3)
```
are **NOT**.

</div>


<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt7

Calculate the mean and median of the set `1, 2, 3, 4, 5, 5, 6, 7, 9, 12, 13`. Don't forget the `c()` format to indicate a list of numbers in `R`!

```{r tryIt7, exercise = TRUE, error = T}
## Replace this line with your code
```

```{r tryIt7-solution}
mean(c(1, 2, 3, 4, 5, 5, 6, 7, 9, 12, 13))
median(c(1, 2, 3, 4, 5, 5, 6, 7, 9, 12, 13))
```
</div>

Also in slide set 1, we were exposed to some graphical summaries. Sure enough, `R` comes equipped with functions to generate these! We'll be returning to these functions throughout the course, but for now let's play around with **histograms.**

To make a histogram out of a single list of numbers in `R`, we use the `hist()` function.

```{r}
hist(c(1.1, 2.1, 2.2, 2.3, 5.5, 7, 12.1))
```

Notice how bad these axis labels and title are! To fix that, we can pass in additional **arguments** into our **call** (this is programming speak for "running" or "executing" a function) to `hist()`. Specifically:

-   The `main` argument controls the title
-   The `xlab` argument controls the _x_-axis label

For example:

```{r}
hist(c(1.1, 2.1, 2.2, 2.3, 5.5, 7, 12.1),
     main = "My First Histogram",
     xlab = "data")
```


<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt8

Generate a histogram from the list of numbers `1, 2, 3, 4, 5, 5, 6, 7, 9, 12, 13`. Title your plot "Histogram of Data", and include an _x_-label that says "values". 

<a style="color:red">**PLEASE NOTE:**</a> There is a known issue with this exercise (related to the server bandwidth, which is out of our control), so **do not try and run the code from within this tutorial.** Instead, write your code in the space below, and then copy-paste it into your Console on the server. \

```{r tryIt8, exercise = TRUE, error = T}
## Replace this line with your code
```

```{r tryIt8-solution}
hist(c(1, 2, 3, 4, 5, 5, 6, 7, 9, 12, 13),
     main = "Histogram of Data",
     xlab = "values")
```
</div>

## What's Next?

Congrats on finishing our  first PSTAT 5LS lab tutorial! Here's what you need to do next.

1)    Navigate to the Lab01 folder on the PSTAT 5LS JupyterHub Server
2)    Make a copy of the file called `lab01_postlab.Rmd`; rename your copy `lab01_NETID_postlab.Rmd` where you replace `NETID` with your NetID (for example, my file would be called `lab01_epmarzban_postlab.Rmd` since my NetID is `epmarzban`)
3)    Fill out your copy of the Lab 01 Post-Lab.

### Next Time

During our next section meeting, we will explore some statistics collected by way of the survey we sent out over the weekend.