---
title: "PSTAT 5LS Lab 05"
author: "Professor Jack Miller, and Ethan Marzban"
date: "Spring 2026"
output: 
   learnr::tutorial:
     pandoc_args: ["--katex"]
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(fontawesome)
library(pstat5lsSBI)
penguins <- read.csv("penguins.csv")
gentoo <- subset(penguins, penguins$species == "Gentoo")
xbar_gentoo <- 217.2352941
sd_gentoo <- 6.585430609
n_gentoo <- 119
cv_gentoo <- 1.980272
```



## Learning Objectives

### `R` Learning Objectives

1)    Create a graphical display of the $t$ distribution.
2)    Generate and interpret R output providing confidence intervals and hypothesis tests for inference for one population mean $\mu$.
3)    Subset a data set to work with a single group.


### Statistical Learning Objectives

1)    Learn to use the $t$ distribution.
2)    Conduct hypothesis tests for one mean and interpret the results
3)    Create confidence intervals for one mean and interpret the results

### Functions and Operations covered in this lab

1)  `plot_t()`
2)  `pt()`
3)  `qt()`
4)  `t.test()`
5)  `subset()`


## Working with Means

Now that we are working with quantitative variables, the parameter we're interested in now is $\mu$ (mu), the population mean. 

The **Central Limit Theorem** for means tells us that, when we collect a "sufficiently large" sample of $n$ independent observations from a population with mean $\mu$ and standard deviation $\sigma$, the sampling distribution of $\bar x$ will be approximately normal with mean $\mu$ and standard error (SE) $\frac{\sigma}{\sqrt{n}}$.

Typically, we will not know the population standard deviation $\sigma$. As soon as we estimate $\sigma$ with the sample standard deviation $s$, we need to use the **_t_ distribution**. The $t$ distribution takes into account the additional variability that we introduce when we estimate $\sigma$.

### The _t_ Distribution

The $t$ distribution, like the standard normal distribution ($N(0, 1)$) that we've seen before, is symmetric about zero and bell-shaped but has "thicker" ("heavier") tails than the normal distribution. 

The $t$ distribution is a "family" of distributions, meaning there are an infinite number of $t$ distributions. We identify which $t$ distribution we're talking about using just one parameter, the **degrees of freedom (df)**.

As the degrees of freedom increase, the $t$ distribution looks more and more like the $N(0, 1)$ distribution.

<img src="https://docs.google.com/drawings/d/e/2PACX-1vQTmZDjtJO5FZAOppuR3o1QCyZlPx8XKTD8OUaTa8qZVOTaZEaUozWqJDqlHycQWD1rpENeDDk-0XXq/pub?w=720">

We can  use the `plot_t()` function in the `pstat5lsSBI` package to make a graphical display of the $t$ distribution. Here are the arguments the `plot_t()` function takes:

- `df`: This is the degrees of freedom for the sample; this week, we can find `df` by computing $n - 1$, where $n$ is the sample size.
- `shadeValues`: This is the value(s) that we wish to identify, and shade either to the left or to the right of. To include two values we will combine them by using the `c()` function.
- `direction`: This is the direction to shade. The choices are: "less", "greater", "beyond", "between". The text must be written in double quotes. 
- `col.shade`: Optional color choice for the graph.


<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt1

Use the `plot_t()` function to plot the _t_ distribution with 3 degrees of freedom, and shade in the area to the right of 1.

```{r tryIt1, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt1-solution}
plot_t(df = 3,
       shadeValues = 1,
       direction = "greater")
```
</div>


### Finding Probabilites Under a _t_ Distribution

We can find probabilities related to the $t$ distribution using the `pt()` function. The `pt()` function is analogous to the `pnorm()` function we used with normal distributions, and admits the following arguments:

- `q`: The observation is sometimes called a **quantile**, thus the **q** notation. 
- `df`: This is the degrees of freedom for the sample; this week, we can find `df` by computing $n - 1$, where $n$ is the sample size.
- `lower.tail`: By default, this argument is set to `TRUE`, meaning that we *want* the lower tail (i.e., to shade to the left). If we *don't want* the lower tail, and we actually want the *upper* tail (i.e., to shade to the right), we should set this to `FALSE`.

The mean of a $t$ distribution is 0, and the standard deviation of a $t$ distribution is a function of the degrees of freedom, so we *don't* need to specify the mean and the standard deviation to use `pt()`.


<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt2

Calculate the area you plotted in the `TryIt1` exercise above. How does this compare to the area to the right of $1$ underneath a standard normal distribution?

```{r tryIt2, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt2-solution}
pt(1, df = 3, lower.tail = FALSE)
## compare this to the ~0.16 we'd expect under the standard normal, by the 68-95-99.7 rule
```

</div>


### Finding the Quantile on a _t_ Distribution

Also, we can use `qt()` to get quantiles of the $t$ distribution. The `qt()` function will be helpful to find $t^*$ critical values needed for confidence intervals for means, and admits the following arguments:

- `p`: The probability to the **left by default** of the quantile we wish to find. If we want the probability to the *right*, we should tinker with `lower.tail` as specified below.
- `df`: This is the degrees of freedom for the sample; this week, we can find `df` by computing $n - 1$, where $n$ is the sample size.
- `lower.tail`: By default, this argument is set to `TRUE`, meaning that we *want* the lower tail (i.e., to shade to the left). If we *don't want* the lower tail, and we actually want the *upper* tail (i.e., to shade to the right), we should set this to `FALSE`.



<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt3

What is the 97.5^th^ percentile (aka quantile) of the _t_(3) distribution (i.e. the _t_ distribution with 3 degrees of freedom)? How does this compare to the 97.5^th^ percentile of the standard normal distribution?

```{r tryIt3, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt3-solution}
qt(0.975, df = 3, lower.tail = TRUE)
## compare this to the 1.96 under the standard normal distribution - quite a bit higher!
```

</div>



## Back to the Penguins!

We've been exposed (several times by now) to the Palmer Penguins dataset, to which we'll now return once again. We've taken the liberty of pre-loading in the data to this lab tutorial, and assigned the result to a variable called `penguins`.

Let's say we want to investigate the population mean flipper length of penguins living in the Palmer Archipelago. In order to use our technology for constructing confidence intervals and performing hypothesis tests for means, we need two conditions to hold. **What are they?**

1)    **Independence:** The observations must be independent of one another. 
2)    **Normality:** The variable follows a normal distribution.

How can we check that we meet both of these conditions?

1)    **Checking Independence:** Verify that we have taken a random sample from the population. If we don't have a random sample, we should consider whether it's reasonable or not to believe that the observations are independent of one another.
2)    **Checking Normality:** Examine a histogram of the sample data. When the sample is small, we require that the sample observations come from a normally distributed population. We can relax this condition more and more for larger and larger sample sizes.

Saying "small $n$" and "large $n$" is still vague, so here are a few guidelines:

- slight skew is okay for sample sizes up to about 15
- moderate skew is okay for sample sizes up to about 30
- strong skew is okay for sample sizes of about 60 or more

Do you think that the penguins data is from a random sample? If not, do you think that the observations of penguins are independent of one another? In fact, as your "hidden instruction" for this week's lab tutorial, indicate your answer to this question in the appropriate place in the Lab 05 Post-Lab document.

**Note:** Checking independence typically means that we need to know how the data was collected. If you're doing your own research, you'll do what you can to make sure of this. For the purposes of PSTAT 5LS, Prof Miller will tell you that the data are from a random sample or that the independence condition is satisfied.

Here's a histogram of the flipper lengths from the Palmer Penguins dataset:

```{r}
hist(penguins$flipper_length_mm,
     main = "Histogram of Flipper Length",
     xlab = "Flipper length (mm)",
     col = "#09847A")
```

That's pretty bimodal... In lecture, Professor Miller mentioned that bimodal distributions are often indicative of subgroups. Sure enough, if we color our histogram based on the species of the individual penguins, the picture becomes a bit clearer:

<img src="https://docs.google.com/drawings/d/e/2PACX-1vTEcgFpev5J8SIfkfc7CRwxOwCHtCvtABTqGVR30rG7YK5Vgm9z1M_qnAR6cH2ZWxkxjpeAhqUaMoCh/pub?w=720">

Adélie and Chinstrap penguins seem to have similar flipper lengths, but Gentoo penguins seem to have noticeably longer flippers. As such, for the purpose of our work in this lab, we will work the flipper lengths for the Gentoo penguins.

### Subsetting the Data

In Lab03, we were introduced to the `which()` function, which gives us the indices of a variable that match a given condition.

<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt4

What are the indices of the rows in the `penguins` dataset that correspond to Gentoo penguins? **Hint:** think about which variable in the dataset might help you here. 

```{r tryIt4, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt4-solution}
which(penguins$species == "Gentoo")
```

</div>

Though we can combine `which()` and indexing (`[]`) just like we did on Lab03, there is another way to subset datasets - using the **`subset()`** function. 

<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt5

Look up the help file for the `subset()` function. Then, extract out only the rows of the `penguins` dataset that correspond to Gentoo penguins, and assign this to a variable called `gentoo`. **Important:** notice that the `gentoo` variable should contain the _actual rows_ of the `penguins` dataset corresponding to Gentoo penguins, not just the _indices_.

```{r tryIt5, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt5-solution}
gentoo <- subset(penguins, penguins$species == "Gentoo")
```

</div>

<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt6

Generate a histogram of the flipper lengths of Gentoo penguins included in the Palmer Penguins Dataset, and describe it using the SOCS mnemonic. 

<a style="color:red"><b>IMPORTANT NOTE</b></a> If your code is not running, this may be due to a server issue. In that case:

1)  Type your code into the box provided below
2)  Copy this code to your Console and run it from there

```{r tryIt6, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt6-solution}
hist(gentoo$flipper_length_mm,
     main = "Histogram of Flipper Length for Gentoo Penguins",
     xlab = "Flipper length (mm)",
     col = "#09847A")

## Shape: Unimodal, with a bit of a right skew
## Outliers: No noticeable outliers
## Center: Around 217 mm, which is the median
## Spread: Around 10 mm, which is the IQR
```

</div>

## Checking Normality

For your convenience, here is the histogram you were supposed to generate in the `TryIt6` exercise in the previous section.

<img src="https://docs.google.com/drawings/d/e/2PACX-1vSkyh4Z9QtaJPMFroqkWZndzbgOaArsRXKy8VWbh351CejY54tvV66_jz6PP47kQvJuyshhLlCrf01D/pub?w=600">

Use the histogram you created to check the normality condition. Does it appear that the flipper lengths for the Gentoo penguins came from a normally distributed population. If not, what can we do?

<details>
<summary>`r fontawesome::fa("lightbulb")` **Show Answers** Click to Expand!</summary>
The histogram of flipper lengths for the Gentoo penguins is unimodal and slightly skewed to the right. Skew does not impact larger data sets as much as it does smaller data sets. Since 119 is a large sample, we can relax the normality condition. 

(The reason we can relax the normality condition is that the Central Limit Theorem tells us that the sampling distribution of the sample mean $\bar{x}$ flipper length will be approximately normal.)
</details>



## Hypothesis Test for a Mean

When we have the data, we can use a new function called `t.test()` to have R compute the confidence interval for the mean flipper length. 

**Note:** We **must have the data to run `t.test()`**. If we only have summary statistics, we need to calculate the test statistic by hand and then use the function `pt()` in R to calculate the p-value. 

According to Australia's Department of Climate Change, Energy, the Environment and Water, Gentoo penguins have an average flipper length of 230 mm. Is the average flipper length for the Gentoo penguins living on Palmer Archipelago different? Use $\alpha = 0.05$.

Thus we are testing: 
$$ \begin{array}{ll}
  H_0:  & \mu = 230 \\
  H_a   & \mu \neq 230
\end{array} $$

where $\mu$ is the mean flipper length in the population of Gentoo penguins living in the Palmer Archipelago.


### Using `t.test()` to Run this Hypothesis Test

Since we have the data in R, let's use `t.test()` to run the hypothesis test.


<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt7

Test the above-specified null and alternative hypotheses ($H_0: \mu = 230$ vs $H_a: \mu \neq 230$) using the `t.test()` function. Make sure to specify the hypothesized mean in the `mu` argument and the alternative in the `alternative` argument (`"two.sided"`, `"less"`, `"greater"`). 

```{r tryIt7, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt7-solution}
t.test(gentoo$flipper_length_mm, 
       mu =  230, 
       alternative =  "two.sided")
```

</div>


When R reports a p-value of `< 2.2e-16`, it is telling us that the p-value is less than $2.2 \times 10^{-16}$. In other words, the p-value is close to zero. Because our p-value is less than the significance level of 0.05, we reject the null hypothesis.

We've decided to reject the null hypothesis, so what's the conclusion?

<details>
<summary>`r fontawesome::fa("lightbulb")` **Show Answers** Click to Expand!</summary>
The data provide convincing evidence that the mean flipper length for the population of Gentoo penguins on the Palmer Archipelago differs from 230 mm.
</details>

\

<div style="background-color:#ede7d8">
### `r fontawesome::fa("triangle-exclamation")` Caution

Don't use language that is too strong when you are writing your conclusions. We are not proving or disproving anything with our analyses. We cannot say that the mean flipper length *is* different from 230 mm. We are just saying that we don't think the mean flipper length is 230 mm because it would have been very unlikely for us to get a sample mean of 217.235 mm in a sample of 119 Gentoo penguins if the mean flipper length in the population of Gentoo penguins is 230 mm.
</div>

## Creating a Confidence Interval for the Mean Flipper Length

Since we rejected the null hypothesis that the mean flipper length of Gentoo penguins is 230 mm, it makes sense to come up with an estimate for this parameter.

Let's create a 95% confidence interval for the mean flipper length of Gentoo penguins.

First, let's do this by hand.

As you learned in lecture, when we have a sample of $n$ independent observations from a nearly normal distribution, a confidence interval for the population mean $\mu$ is

$$\bar x \pm t^{\ast} \cdot \frac{s}{\sqrt{n}}$$
<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt8

Calculate $\bar{x}$, $s$, and $n$, and assign these to variables called `xbar_gentoo`, `sd_gentoo`, and `n_gentoo`, respectively. (It is up to you to remember what these quantities represent in the context of this problem!)

```{r tryIt8, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt8-solution}
## Solutions to this question have been suppressed. 
## Please check with your TA to make sure you have the right answer!
```

</div>

We still need the critical value, $t^{\ast}$. Let's walk through the calculation for finding this in a bit more detail.

### Confidence Level

Suppose we want to construct a confidence interval with a _c_ confidence level (where _c_ is a number between 0 and 1; e.g. a 95\% confidence interval has confidence level _c_ = 0.95). The critical value $t^{\ast}$ is the value such that the area between $-t^{\ast}$ and $+t^{\ast}$ underneath the $t-$distribution with $(n - 1)$ degrees of freedom is _c_. This leads us to the conclusion that the critical value must be the $\left( \frac{1 + c}{2} \right)$^th^ percentile of the $t(n - 1)$ distribution. 

<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt9

Calculate the critical value for a 95\% confidence interval of the population mean flipper length (in millimeters) of Gentoo penguins in the Palmer Archipelago, and assign this to a variable called `cv_gentoo`. 

```{r tryIt9, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt9-solution}
## Solutions to this question have been suppressed. 
## Please check with your TA to make sure you have the right answer!
```

</div>



<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt10

Construct a 95\% confidence interval for the population mean flipper length (in millimeters) of Gentoo penguins in the Palmer Archipelago.

```{r tryIt10, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt10-solution}
## Solutions to this question have been suppressed. 
## Please check with your TA to make sure you have the right answer!
```

</div>

### Interpreting a Confidence Interval for a Population Mean

With any luck, your answer to the `TryIt10` exercise above should be (close to) the interval $(216.0398 \ , \ 218.4308)$. Here's the interpretation of this interval:

> We estimate with 95% confidence that the mean flipper length for all Gentoo penguins in the Palmer Archipelago is between 216.0398 mm and 218.4306 mm.


<div style="background-color:#ede7d8">
### `r fontawesome::fa("triangle-exclamation")` Caution

It is not correct to say: "there is a 95\% chance that the mean flipper length for all Gentoo penguins in the Palmer Archipelago is between 216.0398 mm and 218.4306 mm." This is because: the true value of $\mu$ (the mean flipper length for all Gentoo penguins in the Palmer Archipelago) is a number, and is either in the interval or not. All this interval is doing is providing a range of _plausible_ values for $\mu$.
</div>

### Another Way to Get a Confidence Interval

When we have the data, it's much easier to use `t.test()` to construct a confidence interval. For confidence intervals, we do not specify the `alternative` (we aren't testing hypotheses), but we need to specify the confidence level as a decimal in conf.level. As with hypothesis tests, we can only use `t.test()` if we have the data in R.

Since we only want a confidence interval (we are not doing a hypothesis test), we do not need to specify a hypothesized mean (we don't have one!) nor do we have to specify the alternative. We do have to specify the confidence level.


<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt11

Look up the help file for the `t.test()` function, and use this to generate a 95\% confidence interval for the mean flipper length (in millimeters) of Gentoo penguins in the Palmer Archipelago.

```{r tryIt11, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt11-solution}
t.test(gentoo$flipper_length_mm, conf.level = 0.95)
```

</div>

## What's Next?

Congrats on finishing the Lab05 Tutorial! Here's what you need to do next.

1)    Navigate to the Lab05 folder on the PSTAT 5LS JupyterHub Server
2)    Make a copy of the file called `lab05_postlab.Rmd`; rename your copy `lab05_NETID_postlab.Rmd` where you replace `NETID` with your NetID (for example, my file would be called `lab05_epmarzban_postlab.Rmd` since my NetID is `epmarzban`)
3)    Fill out your copy of the Lab 05 Post-Lab. (Remember: there was a "secret" instruction for your post-lab somewhere in this lab tutorial!)
