---
title: "Lab 05 Post-Lab"
author: "Your Name Here"
date: "Spring 2026"
output: pdf_document
header-includes:
  - \usepackage{mdframed}
urlcolor: purple
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Welcome to the Lab 05 Post-Lab! **Please note:** to encourage further interaction with the associated Lab Tutorial, this Post-Lab will be a little different than previous Post-Labs. Please follow all instructions outlined below fully, as failure to do so may incur point penalties. 

### \textcolor{blue}{Tasks to Complete}

-   Below, type the code you used to generate the confidence interval in the `TryIt10` exercise on the Lab 05 Tutorial. Assume the `xbar_gentoo`, `sd_gentoo`, `n_gentoo`, and `cv_gentoo` variables have already been defined. **Failure to complete this will incur a 0.1 point penalty.**

```{r, eval = FALSE}
Replace this line with your code.
```

\

-   Carry out the secret instruction below. **Failure to complete this will incur a 0.15 point penalty.**

\begin{mdframed}

Replace this line with your text.

\end{mdframed}

\

### Three Things I learned in Lab 05 are:

-   _Replace this line with your text_
-   _Replace this line with your text_
-   _Replace this line with your text_

### Three Questions I have after completing Lab 05 are:

-   _Replace this line with your text_
-   _Replace this line with your text_
-   _Replace this line with your text_

### Three R Functions I learned in Lab 05 are:

-   _Replace this line with your text_
-   _Replace this line with your text_
-   _Replace this line with your text_



## Instructions for Submitting

After filling out the above three sections:

1)    Replace "Your Name Here" at the top of this document with your full name (as it appears on Canvas). 

2)    Click "knit" at the top of your RStudio window (there should be an icon that looks like a ball of yarn). This will turn your document into a PDF

3)    Follow the instructions under the "Downloading a PDF" Section of [this](https://docs.google.com/document/d/12SKM9YPDtStpjVptOvMjVIGZjQcaJFIa_DhJMXE4mMQ/edit?usp=sharing) document to download your PDF to your computer.

4)    Upload the PDF of your work to Gradescope **by 11:59pm on Wednesday May 13, 2026**.

