---
title: "PSTAT 5LS Lab 03"
author: "Professor Jack Miller, and Ethan Marzban"
date: "Spring 2026"
output: 
   learnr::tutorial:
     mathjax: "default"
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(fontawesome)
library(pstat5lsSBI)

set.seed(2573)
sim1 <- simulate_chance_model(chanceSuccess = 0.25, 
                              numDraws = 40, 
                              numRepetitions = 1000)
```



## Learning Objectives

### `R` Learning Objectives

1)    Learn about logical comparisons in R
2)    Learn about reproducible randomness by "setting seeds"
3)    Understand the basics of simulation in R


### Statistical Learning Objectives

1)    Understand how to translate a real data scenario into a simulation setup
2)    Explore sample-to-sample variability by observing a distribution of simulated p-hat values.

### Functions and Operations covered in this lab

1)    Indexing (using `[]`)
2)    `which()`
3)    `length()`
4)    `set.seed()`
5)    `simulate_chance_model()`
6)    `sum()`


## Selecting a Subset of a List

As we saw in the Section Activity last week (Week 2), we sometimes want to select only a smaller portion of a list of numbers. For example, given a list of numbers that represent scores on a hypothetical PSTAT 5LS exam, we may only want to explore scores that were above an 80\%. 

The first thing we need to know is that R assigns an **index** to each element of a list. The index is the position of the element in relation to the leftmost value; that is, the first element has index `1`, the second element has index `2`, etc. 

<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt1

Consider the list of numbers `c(1, 5, 3, 4, 4, 2)`. What would be the index of the number `5`?

```{r tryIt1, exercise = TRUE, error = T}
## DON'T TYPE ANYTHING HERE.
## WHEN YOU ARE READY, CLICK "SOLUTIONS" TO SEE THE ANSWERS
```

```{r tryIt1-solution}
## 2 (since the number "5" is the second number in the list)
```
</div>

If we know the index of a particular element in a list `<list>`, we can extract it using the command
```{r slicing, eval = F}
<list_name>[<index>]
```

For example, 
```{r}
x <- c(5.2, 4.4, 100, 12.14, 302.10, 10.22)
x[4]
```

<details>
<summary>`r fontawesome::fa("magnifying-glass")` **What Is This Code Doing?** Click to Expand!</summary>

-   The first line is assigning a list of numbers to the variable `x`
-   The second line is extracting the **4**^th^ element of `x` (which is `12.14`).

</details>

\

We can select multiple elements by **slicing** (which is what coders sometimes call the act of extracting subsets of lists of elements) along a _list of indices_. For example:

```{r}
x[c(1, 3)]
```


<details>
<summary>`r fontawesome::fa("magnifying-glass")` **What Is This Code Doing?** Click to Expand!</summary>

-   It is extracting the _first_ and _third_ elements of `x` (which we defined above)

</details>

\

<div style="background-color:#ede7d8">
### `r fontawesome::fa("triangle-exclamation")` Caution

Remember: in `R`, to get a list of numbers, don't forget to put the `c` in front! For example, `(1, 2, 3)` is NOT a valid list of numbers in `R`, but `c(1, 2, 3)` is.
</div>

An easy way to get a list of consecutive integers in `R` is to use the `:` operator: `<start>:<stop>` will give you a list of numbers from `<start>` to `<stop>`, inclusive. For example:

```{r}
3:8
```

<div style="background-color:#eaf0e4">
### `r fontawesome::fa("lightbulb")` ProTip
Note that you don't need to put a `c` in front of something generated by the `:` operator.
</div>


<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt2

The following code chunk creates a list of numbers called `my_list`. **Don't worry about what this code - we'll explain it in a future lab.**

```{r}
set.seed(100)
my_list <- rnorm(100)
```

Select the twenty-seventh (27^th^) through thirty-ninth (39^th^) elements, inclusive on both ends, of `my_list`.

```{r tryIt2, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt2-solution}
my_list[27:39]
```
</div>

### Reintroducing `which()`

During our second Section Activity (on Wednesday or Thursday of last week, Week 3, depending on when your Sections meet), we were introduced to the `which()` function. Recall that `which()` returns a list of indices where a given **condition** is true.

For example, to extract the indices of elements of `my_list` that are greater or equal to 1, we can use
```{r}
which(my_list >= 1)
```

R understands most mathematical **comparisons** (e.g. "greater than or equal to"):

| **Name** | **Math Symbol** | **R symbol** |
|:------:|:-------:|:--------:|
| Greater than | $>$ | `>` |
| Less than | $<$ | `<` |
| Greater than or equal to | $\geq$ | `>=` |
| Less than or equal to | $\leq$ | `<=` |
| Equal to | $=$ | `==` |
| Not equal to | $\neq$ | `!=` |


<div style="background-color:#ede7d8">
### `r fontawesome::fa("triangle-exclamation")` Caution

Note that in R, to check whether two things are equal or not, we use a DOUBLE equals sign.
</div>


<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt3

What are the indices of elements of `my_list` that are negative? (**Recall** that zero is not negative.)

```{r tryIt3, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt3-solution}
which(my_list < 0)
```
</div>


\

<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt4

What **proportion** of elements of `my_list` are negative? **Hint:** look up the help file for the `length()` function.

```{r tryIt4, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt4-solution}
length(which(my_list < 0)) / length(my_list)
```
</div>


### Combining Everything

Finally, we can perhaps see how to incorporate `which()` and indexing (using `[]`) to select elements that satisfy a given condition. The general idea is:

-   A call to `which()` gives us the indices of elements that satisfy a given condition
-   Passing the _output_ of `which()` into an indexing operation allows us to extract the actual _elements_ that satisfy a given condition (not just their indices).

For example:
```{r}
which(my_list >= 2)
my_list[which(my_list >= 2)]
```

<details>
<summary>`r fontawesome::fa("magnifying-glass")` **What Is This Code Doing?** Click to Expand!</summary>

-   The first line tells us that the elements with indices 20, 64, and 96 of `my_list` were at least 2
-   The second line tells us that the actual elements of `my_list` that were at least 2 are: 2.310297, 2.581959, and 2.445683

</details>


## Pseudo-Random Numbers

It turns out that humans are very bad at generating numbers randomly. Take a look at this word cloud of results from a survey of students in an introductory statistics course, in which students were asked to provide a random three-digit number. The larger a number is, the more often it showed up in the data.

<img src="https://docs.google.com/drawings/d/e/2PACX-1vRNyts0mYHXqX03yMsmpnKG5kMw87z7o_b4VkP9BJaN-BUA_AwH4HmoguIuyBIG-2b_GNG8atoJVCR3/pub?w=400">

As you can see, people tended to choose 123 a lot, or just typed the same number three times. If these numbers were *truly* random, we'd wouldn't see this -- we'd see most three digit numbers in the data, and each would appear only a few times.

Computers are *much* better at randomness than humans are, but they're actually still quite bad. "Random" numbers generated by computers are meant to *look* and *seem* random, but, the sequences of numbers they generate are completely predictable if you know where that sequence starts. That starting point is called a **seed**.

Think of the seed as, well, a seed: if you know what kind of seed you plant, you know exactly what kind of flowers will grow. By setting a seed, you can predict the exact sequence of "random" numbers that will grow from it.

You can choose the seed that your R session will use with a function called `set.seed()`.

```{r seed}
set.seed(2573)
```

We chose the number 2573 arbitrarily. **Note:** there is no visible output to a call to `set.seed()`.


If you execute the code chunk that contains the `set.seed()` code again, it will start your random number generator at the beginning of the sequence.

When you knit your document, it will also start your random number generator at the beginning of the sequence.

### Simulating a Chance Model

In lecture, we are learning about how to simulate results using a chance model (that assumes the null hypothesis is true) to evaluate the probability of getting results like we saw in the data or something even more extreme.

We will be using a7 special package called `pstat5lsSBI` that is available for install through GitHub. (SBI stands for "simulation-based inference.") Our lab projects will have this package pre-installed; you can find instructions on how to install it for a new RStudio project you start in JupyterHub or on your personal RStudio installation on Canvas.

Let's take a look at our first SBI function, `simulate_chance_model()`.

#### `simulate_chance_model()`

The function admits the following arguments:

-   `chanceSuccess`: a number between 0 and 1 representing the probability of observing a "success" 
-   `numDraws`: the number of times to draw a poker chip from the bag needed to complete one repetition of the simulation (this will be equal to the sample size)
-   `numRepetitions`: the number of times to repeat the simulation process (this will be an arbitrary large number so that we can see what the chance model looks like)

Think of a "success" as answering yes to a particular question. For example, a "success" in the Doris and Buzz example occurs when Buzz pushes the correct button.  

We can now apply this to an example!

Before we do, just to make sure you're paying attention - at the very top of your post-lab, please write the name of your favorite fruit! (There will be a 0.25 point deduction for failing to do so.)


## Simulation Example: Setting Up the Simulation

Lee is a teacher at a local high school who wanted to assess whether or not dogs physically resemble their owners enough for people to be able to correctly match a dog to their owner better than if just guessing. Lee, who is also a dog owner, showed pictures of four dogs to her class of 40 students. One photo was of the teacher's dog (Yoda) and the other three photos were of dogs belonging to Lee's friends. The students were asked to guess which dog was actually the teacher's. If dogs do not physically resemble their owners, the students would get a correct match 25\% of the time, since the students would be equally likely to choose any of the four dogs. It turned out that 17 of the 40 students correctly picked out the teacher's dog.

Let's go through the entire procedure from start to finish.


<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt5

What is the observed (sample) proportion of correct guesses? Assign the correct answer to a variable called `p_hat`.

```{r tryIt5, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt5-solution}
p_hat <- 17/40
```
</div>

<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt6

What are the hypotheses to be tested? State the hypotheses using symbols. Be sure to define the parameter. You'll want to define both $H_0$ and $H_A$.

where *write your symbol here* represents the *write your answer here*.

<details>
<summary>`r fontawesome::fa("lightbulb")` **Answers** (Click to Expand)</summary>

-   $H_0: p = 0.25$
-   $H_A: p > 0.25$

where $p$ represents the **proportion** of students who **correctly** guess which dog belongs to their teacher.

</details>

</div>


### Poker Chip Analogy

In lecture, we were introduced to an analogy for thinking about randomization studies (like the one we're conducting in this lab tutorial right now!) The basic idea is: use a blue poker chip to represent a "success" and a yellow poker chip to represent a "failure". 

Under the null hypothesis, there is a 25\% chance that a particular student correctly identifies Lee's dog as Yoda. As such, we can imagine "correctly matching a dog with their owner" is a random event that (assuming the null is true) occurs with 25\% probability. This means that the chance of a student _incorrectly_ matching a dog with its owner is 75\%. 


<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt6

-   We have to draw one poker chip from the bag. What does this action represent in terms of the context of our hypothesis test?

-   What does the blue poker chip represent in terms of the context of our hypothesis test?

-   What does the yellow poker chip represent in terms of the context of our hypothesis test?

-   What is the chance of drawing a blue poker chip from the bag?

-   What is the number of times we should draw poker chips from the bag in order to replicate this study one time?


<details>
<summary>`r fontawesome::fa("lightbulb")` **Answers** (Click to Expand)</summary>


-   Drawing one chip represents the outcome of one particular student guessing which dog is their teacher's

-   A blue poker chip represents a student _correctly_ matching their teacher's dog to their teacher

-   A yellow poker chip represents a student _incorrectly_ matching their teacher's dog to their teacher

-   Under the null hypothesis, there is a 25\% chance that a student correctly matches their teacher's dog with their teacher; hence, 25\% of our chips should be blue.

-   The class size is 40, meaning one sample should be of size 40.


</details>

</div>

<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt7

Fill in the following table on a separate sheet of paper. (You don't need to turn this in, but you should try it on your own as we may ask you questions like these on exams!)

|                   | **Assuming the chance model...** |
|-------------------|----------------------------------|
| One draw          | *Write your answer here*         |
| Blue poker chip   | *Write your answer here*         |
| Yellow poker chip | *Write your answer here*         |
| Chance of blue    | *Write your answer here*         |
| One repetition    | *Write your answer here*         |

<details>
<summary>`r fontawesome::fa("lightbulb")` **Answers** (Click to Expand)</summary>

|                   | **Assuming the chance model...** |
|-------------------|----------------------------------|
|          One draw | *one student guessing which dog is their teacher's* |
|   Blue poker chip | *the student guesses the dog correctly* |
| Yellow poker chip | *the student has an incorrect guess* |
|    Chance of blue | *0.25* |
|    One repetition | *40 draws (to represent the 40 students)* |

</details>

</div>


## Simulation Example: Conducting the Simulation


### Running the Simulation 

Let's try this out for a small number of repetitions, say, 10, so that we can see what the output is from the function.

```{r simulate1, echo = TRUE}
simulate_chance_model(chanceSuccess = 0.25, 
                      numDraws = 40, 
                      numRepetitions = 10)
```

<details>
<summary>`r fontawesome::fa("magnifying-glass")` **What Is This Code Doing?** Click to Expand!</summary>

This code is running our chance model 10 times. Each outputted value represents the proportion of students in a hypothetical class of size 40 correctly identified Lee's dog as Yoda.

</details>

<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt8

Run the simulation 1000 times, and assign the result to a variable called `sim1`.

```{r tryIt8, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt8-solution}
sim1 <- simulate_chance_model(chanceSuccess = 0.25, 
                              numDraws = 40, 
                              numRepetitions = 1000)
```
</div>


### Visualizing the Results

Of course, we'd like some way to visualize the result of our simulation. Given that `sim1` represents a list of 1000 numbers (each representing the proportion of students in a hypothetical class of size 40 that correctly identified Lee's dog as Yoda), a natural thing to do is: generate a histogram!


<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt9

Generate a histogram of the `sim1` values. Make sure the x-axis values range on your plot range from 0 to 1. **Hint:** Look up the help file for `hist()` and pay attention to the `xlim` argument. \ 

<a style="color:red"><b>IMPORTANT NOTE</b></a> If your code is not running, this may be due to a server issue. In that case:

1)  Type your code into the box provided below
2)  Copy this code to your Console and run it from there

```{r tryIt9, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt9-solution}
hist(sim1,
     main = "Histogram of 1000 Simulation Results",
     xlab = "Simulated Proportion of Correct Guesses",
     xlim = c(0, 1))
```
</div>

For posterity's sake, here is an image of the histogram you should see. We've included a red line at $17/40$, which was the proportion observed in Lee's classroom.

<img src="https://docs.google.com/drawings/d/e/2PACX-1vQEA7fre_gQ4iV9VsvZZ7kd89-bPZIIFeB8XBVlU2r2wktk7vjllKwi9IrsId1NOGf9sALMOZYTxLQp/pub?w=480&amp;h=360">

### Finding the Probability of Getting This or More Extreme

As we learned in lecture, we want to calculate the *probability that we will get our observed* proportion or a result that is *even more extreme* than what we observed.

Since $H_A: p > 0.25$, the "as or more extreme" is actually **greater than or equal to** $\hat{p}$ of $17/40 = 0.425$.

Let $A$ represent something of interest to us. In this example, an observation in $A$ is a simulated proportion of 17/40 or greater. The number of observations in $A$ is the sum of the simulated proportions that are 17/40, 18/40, ..., or 40/40 To find the probability of $A$ in general, we use this formula: $$P(A) = \frac{\text{number of observations of A}}{\text{total observations}}$$ \

So we will be:

1.  looking for the number of observations that have a simulated proportion of 17/40 **or greater**
2.  divide this number of observations by **1000**, the number of times we ran the simulation



<div style="background-color:#dfedf7">
### `r fontawesome::fa("laptop-code")` TryIt10

Calculate the _p_-value. **Hint:** remember `which()` and `length()`.

```{r tryIt10, exercise = TRUE, error = T}
## Replace this line with your answers
```

```{r tryIt10-solution}
length(which(sim1 >= 17/40)) / 1000
```
</div>


### Conclusion

What do the results tell us about our research question?

Do we have enough evidence to support the claim that dogs physically resemble their owners enough for people to be able to correctly match a dog to their owner **better** than if just guessing?

Our estimated p-value of 0.006 is **quite rare**. We will reject the null hypothesis at any the commonly-used $\alpha = 0.05$ significance level.

We **do** have enough evidence to support the claim that dogs physically resemble their owners enough for people to correctly match a dog to their owner better than if just guessing.

## What's Next?

Congrats on finishing the Lab03 Tutorial! Here's what you need to do next.

1)    Navigate to the Lab03 folder on the PSTAT 5LS JupyterHub Server
2)    Make a copy of the file called `lab03_postlab.Rmd`; rename your copy `lab03_NETID_postlab.Rmd` where you replace `NETID` with your NetID (for example, my file would be called `lab03_epmarzban_postlab.Rmd` since my NetID is `epmarzban`)
3)    Fill out your copy of the Lab 03 Post-Lab. (Remember: there was a "secret" instruction for your post-lab somewhere in this lab tutorial!)

### Next Time

During our next section meeting, we will ...